---
title: "SDTM_oak"
author: "Jie He"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reference: 

SDTM programming in R using {sdtm.oak} package: 
https://www.youtube.com/watch?v=H0FdhG9_ttU


Resources mentioned in the workshop:

Workshop GitHub Repository https://github.com/pharmaverse/rinpharma-2024-SDTM-workshop
{sdtm.oak} SDTM transformation engine https://pharmaverse.github.io/sdtm.oak/
CDISC Oak Garden https://wiki.cdisc.org/display/oakgarden
Oak Garden Slack channel https://oakgarden.slack.com/
How to learn R as a SAS user https://posit.co/blog/how-to-learn-r-as-a-sas-user/


Notes:

Should know basics of CDISC standards (SDTM domains)

GPT-4o summary of CDISC standards using search function that I reviewed and believed to be accurate are listed as follows:

1) CDISC SDTM Standards: https://www.cdisc.org/standards/foundational/sdtm Provides comprehensive information on SDTM, including the latest versions and implementation guides.
2) SDTM Implementation Guide (SDTMIG): https://www.cdisc.org/standards/foundational/sdtmig Offers detailed guidance on the structure and content of each domain, including examples and best practices. 
3) !!READ THIS FIRST!! Quanticate's Guide to CDISC SDTM Standards and Domains: https://www.quanticate.com/blog/bid/51830/cdisc-sdtm-v3-1-2-theory-and-application An overview of SDTM theory and application in clinical trials.

- SDTM inputs raw data while ADaM inputs SDTM datasets with well-defined structure.
- SDTM.oak has algorithms ready to transform raw data into target SDTM data model. List of algorithms and descriptions: https://pharmaverse.github.io/sdtm.oak/articles/algorithms.html#list-of-algorithms . 


# Set up workspace
```{r}
# install.packages("sdtm.oak", dependencies = TRUE)
# install.packages("admiral", dependencies = TRUE)

library(sdtm.oak)
library(admiral)
library(dplyr)
```
<!-- Create CM Domain -->

# Read CT Specification and raw data
```{r}
study_ct <- read.csv(file="../datasets/sdtm_ct.csv")

cm_raw <- read.csv(file="../datasets/cm_raw_data_cdash.csv", stringsAsFactors = FALSE) 
```

# Convert Blank Strings Into NAs
```{r}
cm_raw <- admiral::convert_blanks_to_na(cm_raw)
```

# This code automatically generates oak_id
```{r}
cm_raw <- cm_raw %>%
  generate_oak_id_vars(
    pat_var = "PATNUM",
    raw_src = "cm_raw"
  )
```

# Read in DM domain to derive study day
```{r}
dm <- read.csv("../datasets/dm.csv")

dm <- admiral::convert_blanks_to_na(dm)
```

# Create CM domain. The first step in creating CM domain is to create the topic variable
- CM_PDF
- VITALSIGNS_eCRF
```{r}
cm <-
  # Derive topic variable
  # Map CMTRT using assign_no_ct, raw_var=IT.CMTRT,tgt_var=CMTRT
  assign_no_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMTRT",
    tgt_var = "CMTRT"
  ) %>%
  # Map CMINDC using assign_no_ct, raw_var=IT.CMINDC,tgt_var=CMINDC
  assign_no_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMINDC",
    tgt_var = "CMINDC",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSTXT using condition_add and assign_no_ct, raw_var=IT.CMDSTXT,tgt_var=CMDOS
  # If IT.CMDSTXT is numeric, map it to CMDOS
  assign_no_ct(
    raw_dat = condition_add(cm_raw, grepl("^-?\\d*(\\.\\d+)?(e[+-]?\\d+)?$", cm_raw$IT.CMDSTXT)),
    raw_var = "IT.CMDSTXT",
    tgt_var = "CMDOS",
    id_vars = oak_id_vars()
  ) %>%
  # Map qualifier CMDOSTXT using condition_add & assign_no_ct, raw_var=IT.CMDSTXT,tgt_var=CMDOSTXT
  # If IT.CMDSTXT is character, map it to CMDOSTXT
  assign_no_ct(
    raw_dat = condition_add(cm_raw, grepl("[^0-9eE.-]", cm_raw$IT.CMDSTXT)),
    raw_var = "IT.CMDSTXT",
    tgt_var = "CMDOSTXT",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSU and apply CT using assign_ct, raw_var=IT.CMDOSU,tgt_var=CMDOSU
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMDOSU",
    tgt_var = "CMDOSU",
    ct_spec = study_ct,
    ct_clst = "C71620",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSFRM and apply CT using assign_ct, raw_var=IT.CMDOSFRM,tgt_var=CMDOSFRM
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMDOSFRM",
    tgt_var = "CMDOSFRM",
    ct_spec = study_ct,
    ct_clst = "C66726",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSFRQ using assign_ct, raw_var=IT.CMDOSFRQ,tgt_var=CMDOSFRQ
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMDOSFRQ",
    tgt_var = "CMDOSFRQ",
    ct_spec = study_ct,
    ct_clst = "C71113",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMROUTE using assign_ct, raw_var=IT.CMROUTE,tgt_var=CMROUTE
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMROUTE",
    tgt_var = "CMROUTE",
    ct_spec = study_ct,
    ct_clst = "C66729",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMSTDTC using assign_no_ct, raw_var=IT.CMSTDAT,tgt_var=CMSTDTC
  assign_datetime(
    raw_dat = cm_raw,
    raw_var = "IT.CMSTDAT",
    tgt_var = "CMSTDTC",
    raw_fmt = c("d-m-y"),
    raw_unk = c("UN", "UNK")
  ) %>%
  # Map CMENRTPT using assign_ct, raw_var=IT.CMONGO,tgt_var=CMENRTPT
  # If IT.CMONGO is Yes then CM.CMENRTPT = 'ONGOING'
  hardcode_ct(
    raw_dat = condition_add(cm_raw, IT.CMONGO == "Yes"),
    raw_var = "IT.CMONGO",
    tgt_var = "CMENRTPT",
    ct_spec = study_ct,
    ct_clst = "C66728",
    tgt_val = "Ongoing",
    id_vars = oak_id_vars()
  ) %>%
  # Map CM.CMENTPT using hardcode_no_ct, raw_var = IT.CMONGO, tgt_var=CMMENTPT
  # If IT.CMONGO is Yes then CM.CMENTPT = 'DATE OF LAST ASSESSMENT'
  hardcode_no_ct(
    raw_dat =  condition_add(cm_raw, IT.CMONGO == "Yes"),
    raw_var = "IT.CMONGO",
    tgt_var = "CMENTPT",
    tgt_val = "DATE OF LAST ASSESSMENT",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMENDTC using assign_no_ct, raw_var=IT.CMENDAT,tgt_var=CMENDTC
  assign_datetime(
    raw_dat = cm_raw,
    raw_var = "IT.CMENDAT",
    tgt_var = "CMENDTC",
    raw_fmt = c("d-m-y"),
    raw_unk = c("UN", "UNK")
  ) %>%
  dplyr::mutate(
    STUDYID = "test_study",
    DOMAIN = "CM",
    CMCAT = "GENERAL CONMED",
    USUBJID = paste0("test_study", "-", cm_raw$PATNUM)
  ) %>%
  derive_seq(tgt_var = "CMSEQ",
             rec_vars= c("USUBJID", "CMTRT")) %>%
  derive_study_day(
    sdtm_in = .,
    dm_domain = dm,
    tgdt = "CMENDTC",
    refdt = "RFXSTDTC",
    study_day_var = "CMENDY"
  ) %>%
  derive_study_day(
    sdtm_in = .,
    dm_domain = dm,
    tgdt = "CMSTDTC",
    refdt = "RFXSTDTC",
    study_day_var = "CMSTDY"
  ) %>%
  dplyr::select("STUDYID", "DOMAIN", "USUBJID", "CMSEQ", "CMTRT", "CMCAT", "CMINDC", 
                "CMDOS", "CMDOSTXT", "CMDOSU", "CMDOSFRM", "CMDOSFRQ", "CMROUTE", 
                "CMSTDTC", "CMENDTC","CMSTDY", "CMENDY", "CMENRTPT", "CMENTPT")
```


# Read in raw data for vitals
```{r}
vitals_raw <- read.csv("../datasets/vitals_raw_data.csv", 
                   stringsAsFactors = FALSE) 

vitals_raw <- admiral::convert_blanks_to_na(vitals_raw)
```


# derive oak_id_vars
```{r}
vitals_raw <- vitals_raw %>%
  generate_oak_id_vars(
    pat_var = "PATNUM",
    raw_src = "vitals_raw"
  )
```

<!-- Create VS domain. -->

Create the topic variable and corresponding qualifiers for the VS domain.

very efficient way of mapping SYS_BP and DIA_BP in raw_vitals with qualifiers which were then stored in new datasets.

# Map topic variable SYSBP and its qualifiers.
```{r}
vs_sysbp <-
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "SYS_BP",
    tgt_var = "VSTESTCD",
    tgt_val = "SYSBP",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  # Filter for records where VSTESTCD is not empty.
  # Only these records need qualifier mappings.
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "SYS_BP",
    tgt_var = "VSTEST",
    tgt_val = "Systolic Blood Pressure",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRES using assign_no_ct algorithm
  assign_no_ct(
    raw_dat = vitals_raw,
    raw_var = "SYS_BP",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRESU using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "SYS_BP",
    tgt_var = "VSORRESU",
    tgt_val = "mmHg",
    ct_spec = study_ct,
    ct_clst = "C66770",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSPOS using assign_ct algorithm
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "SUBPOS",
    tgt_var = "VSPOS",
    ct_spec = study_ct,
    ct_clst = "C71148",
    id_vars = oak_id_vars()
  )
```


# Map topic variable DIABP and its qualifiers.
```{r}
vs_diabp <-
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "DIA_BP",
    tgt_var = "VSTESTCD",
    tgt_val = "DIABP",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "DIA_BP",
    tgt_var = "VSTEST",
    tgt_val = "Diastolic Blood Pressure",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRES using assign_no_ct algorithm
  assign_no_ct(
    raw_dat = vitals_raw,
    raw_var = "DIA_BP",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRESU using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "DIA_BP",
    tgt_var = "VSORRESU",
    tgt_val = "mmHg",
    ct_spec = study_ct,
    ct_clst = "C66770",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSPOS using assign_ct algorithm
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "SUBPOS",
    tgt_var = "VSPOS",
    ct_spec = study_ct,
    ct_clst = "C71148",
    id_vars = oak_id_vars()
  )
```


# Map topic variable PULSE and its qualifiers.
```{r}
vs_pulse <-
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "PULSE",
    tgt_var = "VSTESTCD",
    tgt_val = "PULSE",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "PULSE",
    tgt_var = "VSTEST",
    tgt_val = "Pulse Rate",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRES using assign_no_ct algorithm
  assign_no_ct(
    raw_dat = vitals_raw,
    raw_var = "PULSE",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRESU using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "PULSE",
    tgt_var = "VSORRESU",
    tgt_val = "beats/min",
    ct_spec = study_ct,
    ct_clst = "C66770",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSPOS using assign_ct algorithm
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "SUBPOS",
    tgt_var = "VSPOS",
    ct_spec = study_ct,
    ct_clst = "C71148",
    id_vars = oak_id_vars()
  )
```

# Map topic variable RESP from the raw variable RESPRT and its qualifiers.
```{R}
vs_resp <-
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "RESPRT",
    tgt_var = "VSTESTCD",
    tgt_val = "RESP",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "RESPRT",
    tgt_var = "VSTEST",
    tgt_val = "Respiratory Rate",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRES using assign_no_ct algorithm
  assign_no_ct(
    raw_dat = vitals_raw,
    raw_var = "RESPRT",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRESU using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "RESPRT",
    tgt_var = "VSORRESU",
    tgt_val = "breaths/min",
    ct_spec = study_ct,
    ct_clst = "C66770",
    id_vars = oak_id_vars()
  )
```

# Map topic variable TEMP from raw variable TEMP and its qualifiers.
```{r}
vs_temp <-
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "TEMP",
    tgt_var = "VSTESTCD",
    tgt_val = "TEMP",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "TEMP",
    tgt_var = "VSTEST",
    tgt_val = "Temperature",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRES using assign_no_ct algorithm
  assign_no_ct(
    raw_dat = vitals_raw,
    raw_var = "TEMP",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRESU using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "TEMP",
    tgt_var = "VSORRESU",
    tgt_val = "C",
    ct_spec = study_ct,
    ct_clst = "C66770",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSLOC from TEMPLOC using assign_ct
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "TEMPLOC",
    tgt_var = "VSLOC",
    ct_spec = study_ct,
    ct_clst = "C74456",
    id_vars = oak_id_vars()
  )
```


# Map topic variable OXYSAT from raw variable OXY_SAT and its qualifiers.
```{r}
vs_oxysat <-
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "OXY_SAT",
    tgt_var = "VSTESTCD",
    tgt_val = "OXYSAT",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "OXY_SAT",
    tgt_var = "VSTEST",
    tgt_val = "Oxygen Saturation",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRES using assign_no_ct algorithm
  assign_no_ct(
    raw_dat = vitals_raw,
    raw_var = "OXY_SAT",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSORRESU using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "OXY_SAT",
    tgt_var = "VSORRESU",
    tgt_val = "%",
    ct_spec = study_ct,
    ct_clst = "C66770",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSLAT using assign_ct from raw variable LAT
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "LAT",
    tgt_var = "VSLAT",
    ct_spec = study_ct,
    ct_clst = "C99073",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSLOC using assign_ct from raw variable LOC
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "LOC",
    tgt_var = "VSLOC",
    ct_spec = study_ct,
    ct_clst = "C74456",
    id_vars = oak_id_vars()
  )
```


# Map topic variable VSALL from raw variable ASMNTDN with the logic if ASMNTDN  == Yes then VSTESTCD = VSALL

more complex mapping with condition statements
```{r}
vs_vsall <-
  hardcode_ct(
    raw_dat = condition_add(vitals_raw, ASMNTDN == "Yes"),
    raw_var = "ASMNTDN",
    tgt_var = "VSTESTCD",
    tgt_val = "VSALL",
    ct_spec = study_ct,
    ct_clst = "C66741"
  ) %>%
  dplyr::filter(!is.na(.data$VSTESTCD)) %>%
  # Map VSTEST using hardcode_ct algorithm
  hardcode_ct(
    raw_dat = vitals_raw,
    raw_var = "ASMNTDN",
    tgt_var = "VSTEST",
    tgt_val = "Vital Signs",
    ct_spec = study_ct,
    ct_clst = "C67153",
    id_vars = oak_id_vars()
  )

# Combine all the topic variables into a single data frame.
vs_combined <- dplyr::bind_rows(
  vs_vsall, vs_sysbp, vs_diabp, vs_pulse, vs_resp,
  vs_temp, vs_oxysat
)
```



# Map qualifiers common to all topic variables
```{r}
vs <- vs_combined %>%
  # Map VSDTC using assign_ct algorithm
  assign_datetime(
    raw_dat = vitals_raw,
    raw_var = c("VTLD", "VTLTM"),
    tgt_var = "VSDTC",
    raw_fmt = c(list(c("d-m-y", "dd-mmm-yyyy")), "H:M")
  ) %>%
  # Map VSTPT from TMPTC using assign_ct
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "TMPTC",
    tgt_var = "VSTPT",
    ct_spec = study_ct,
    ct_clst = "TPT",
    id_vars = oak_id_vars()
  ) %>%
  # Map VSTPTNUM from TMPTC using assign_ct
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "TMPTC",
    tgt_var = "VSTPTNUM",
    ct_spec = study_ct,
    ct_clst = "TPTNUM",
    id_vars = oak_id_vars()
  ) %>%
  # Map VISIT from VISIT_NAME using assign_ct
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "VISIT_NAME",
    tgt_var = "VISIT",
    ct_spec = study_ct,
    ct_clst = "VISIT",
    id_vars = oak_id_vars()
  ) %>%
  # Map VISITNUM from VISIT_NAME using assign_ct
  assign_ct(
    raw_dat = vitals_raw,
    raw_var = "VISIT_NAME",
    tgt_var = "VISITNUM",
    ct_spec = study_ct,
    ct_clst = "VISITNUM",
    id_vars = oak_id_vars()
  ) %>%
  dplyr::mutate(
    STUDYID = "test_study",
    DOMAIN = "VS",
    VSCAT = "VITAL SIGNS",
    USUBJID = paste0("test_study", "-", .data$patient_number)
  ) %>%
  derive_seq(tgt_var = "VSSEQ",
             rec_vars= c("USUBJID", "VISITNUM", "VSTPTNUM", "VSTESTCD")) %>%
  # A bug in derive_study_day V0.1 that clears the time values in VSDTC
  derive_study_day(
    sdtm_in = .,
    dm_domain = dm,
    tgdt = "VSDTC",
    refdt = "RFXSTDTC",
    study_day_var = "VSDY"
  ) %>%
  dplyr::select("STUDYID", "DOMAIN", "USUBJID", "VSSEQ",
                "VSTESTCD", "VSTEST", "VSCAT", "VSPOS", 
                "VSORRES", "VSORRESU", "VSLOC", "VSLAT", 
                "VISIT", "VISITNUM", "VSDY", "VSTPT", "VSTPTNUM", "VSDTC" )
```



