---
title: "SDTM_oak"
author: "Jie He"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reference: 

SDTM programming in R using {sdtm.oak} package: 
https://www.youtube.com/watch?v=H0FdhG9_ttU


Resources mentioned in the workshop:

Workshop GitHub Repository https://github.com/pharmaverse/rinpharma-2024-SDTM-workshop
{sdtm.oak} SDTM transformation engine https://pharmaverse.github.io/sdtm.oak/
CDISC Oak Garden https://wiki.cdisc.org/display/oakgarden
Oak Garden Slack channel https://oakgarden.slack.com/
How to learn R as a SAS user https://posit.co/blog/how-to-learn-r-as-a-sas-user/


Notes:

Should know basics of CDISC standards (SDTM domains)

GPT-4o summary of CDISC standards using search function that I reviewed and believed to be accurate are listed as follows:

1) CDISC SDTM Standards: https://www.cdisc.org/standards/foundational/sdtm Provides comprehensive information on SDTM, including the latest versions and implementation guides.
2) SDTM Implementation Guide (SDTMIG): https://www.cdisc.org/standards/foundational/sdtmig Offers detailed guidance on the structure and content of each domain, including examples and best practices. 
3) !!READ THIS FIRST!! Quanticate's Guide to CDISC SDTM Standards and Domains: https://www.quanticate.com/blog/bid/51830/cdisc-sdtm-v3-1-2-theory-and-application An overview of SDTM theory and application in clinical trials.

- SDTM inputs raw data while ADaM inputs SDTM datasets with well-defined structure.
- SDTM.oak has algorithms ready to transform raw data into target SDTM data model. List of algorithms and descriptions: https://pharmaverse.github.io/sdtm.oak/articles/algorithms.html#list-of-algorithms . 


# Set up workspace
```{r}
# install.packages("sdtm.oak", dependencies = TRUE)
# install.packages("admiral", dependencies = TRUE)

library(sdtm.oak)
library(admiral)
library(dplyr)
```
# Read CT Specification and raw data
```{r}
study_ct <- read.csv(file="../datasets/sdtm_ct.csv")

cm_raw <- read.csv(file="../datasets/cm_raw_data_cdash.csv", stringsAsFactors = FALSE) 
```

# Convert Blank Strings Into NAs
```{r}
cm_raw <- admiral::convert_blanks_to_na(cm_raw)
```

# This code automatically generates oak_id
```{r}
cm_raw <- cm_raw %>%
  generate_oak_id_vars(
    pat_var = "PATNUM",
    raw_src = "cm_raw"
  )
```

# Read in DM domain to derive study day
```{r}
dm <- read.csv("../datasets/dm.csv")

dm <- admiral::convert_blanks_to_na(dm)
```

# Create CM domain. The first step in creating CM domain is to create the topic variable
- CM_PDF
- VITALSIGNS_eCRF
```{r}
cm <-
  # Derive topic variable
  # Map CMTRT using assign_no_ct, raw_var=IT.CMTRT,tgt_var=CMTRT
  assign_no_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMTRT",
    tgt_var = "CMTRT"
  ) %>%
  # Map CMINDC using assign_no_ct, raw_var=IT.CMINDC,tgt_var=CMINDC
  assign_no_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMINDC",
    tgt_var = "CMINDC",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSTXT using condition_add and assign_no_ct, raw_var=IT.CMDSTXT,tgt_var=CMDOS
  # If IT.CMDSTXT is numeric, map it to CMDOS
  assign_no_ct(
    raw_dat = condition_add(cm_raw, grepl("^-?\\d*(\\.\\d+)?(e[+-]?\\d+)?$", cm_raw$IT.CMDSTXT)),
    raw_var = "IT.CMDSTXT",
    tgt_var = "CMDOS",
    id_vars = oak_id_vars()
  ) %>%
  # Map qualifier CMDOSTXT using condition_add & assign_no_ct, raw_var=IT.CMDSTXT,tgt_var=CMDOSTXT
  # If IT.CMDSTXT is character, map it to CMDOSTXT
  assign_no_ct(
    raw_dat = condition_add(cm_raw, grepl("[^0-9eE.-]", cm_raw$IT.CMDSTXT)),
    raw_var = "IT.CMDSTXT",
    tgt_var = "CMDOSTXT",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSU and apply CT using assign_ct, raw_var=IT.CMDOSU,tgt_var=CMDOSU
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMDOSU",
    tgt_var = "CMDOSU",
    ct_spec = study_ct,
    ct_clst = "C71620",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSFRM and apply CT using assign_ct, raw_var=IT.CMDOSFRM,tgt_var=CMDOSFRM
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMDOSFRM",
    tgt_var = "CMDOSFRM",
    ct_spec = study_ct,
    ct_clst = "C66726",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMDOSFRQ using assign_ct, raw_var=IT.CMDOSFRQ,tgt_var=CMDOSFRQ
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMDOSFRQ",
    tgt_var = "CMDOSFRQ",
    ct_spec = study_ct,
    ct_clst = "C71113",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMROUTE using assign_ct, raw_var=IT.CMROUTE,tgt_var=CMROUTE
  assign_ct(
    raw_dat = cm_raw,
    raw_var = "IT.CMROUTE",
    tgt_var = "CMROUTE",
    ct_spec = study_ct,
    ct_clst = "C66729",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMSTDTC using assign_no_ct, raw_var=IT.CMSTDAT,tgt_var=CMSTDTC
  assign_datetime(
    raw_dat = cm_raw,
    raw_var = "IT.CMSTDAT",
    tgt_var = "CMSTDTC",
    raw_fmt = c("d-m-y"),
    raw_unk = c("UN", "UNK")
  ) %>%
  # Map CMENRTPT using assign_ct, raw_var=IT.CMONGO,tgt_var=CMENRTPT
  # If IT.CMONGO is Yes then CM.CMENRTPT = 'ONGOING'
  hardcode_ct(
    raw_dat = condition_add(cm_raw, IT.CMONGO == "Yes"),
    raw_var = "IT.CMONGO",
    tgt_var = "CMENRTPT",
    ct_spec = study_ct,
    ct_clst = "C66728",
    tgt_val = "Ongoing",
    id_vars = oak_id_vars()
  ) %>%
  # Map CM.CMENTPT using hardcode_no_ct, raw_var = IT.CMONGO, tgt_var=CMMENTPT
  # If IT.CMONGO is Yes then CM.CMENTPT = 'DATE OF LAST ASSESSMENT'
  hardcode_no_ct(
    raw_dat =  condition_add(cm_raw, IT.CMONGO == "Yes"),
    raw_var = "IT.CMONGO",
    tgt_var = "CMENTPT",
    tgt_val = "DATE OF LAST ASSESSMENT",
    id_vars = oak_id_vars()
  ) %>%
  # Map CMENDTC using assign_no_ct, raw_var=IT.CMENDAT,tgt_var=CMENDTC
  assign_datetime(
    raw_dat = cm_raw,
    raw_var = "IT.CMENDAT",
    tgt_var = "CMENDTC",
    raw_fmt = c("d-m-y"),
    raw_unk = c("UN", "UNK")
  ) %>%
  dplyr::mutate(
    STUDYID = "test_study",
    DOMAIN = "CM",
    CMCAT = "GENERAL CONMED",
    USUBJID = paste0("test_study", "-", cm_raw$PATNUM)
  ) %>%
  derive_seq(tgt_var = "CMSEQ",
             rec_vars= c("USUBJID", "CMTRT")) %>%
  derive_study_day(
    sdtm_in = .,
    dm_domain = dm,
    tgdt = "CMENDTC",
    refdt = "RFXSTDTC",
    study_day_var = "CMENDY"
  ) %>%
  derive_study_day(
    sdtm_in = .,
    dm_domain = dm,
    tgdt = "CMSTDTC",
    refdt = "RFXSTDTC",
    study_day_var = "CMSTDY"
  ) %>%
  dplyr::select("STUDYID", "DOMAIN", "USUBJID", "CMSEQ", "CMTRT", "CMCAT", "CMINDC", 
                "CMDOS", "CMDOSTXT", "CMDOSU", "CMDOSFRM", "CMDOSFRQ", "CMROUTE", 
                "CMSTDTC", "CMENDTC","CMSTDY", "CMENDY", "CMENRTPT", "CMENTPT")
```

